---
title: "Cohort/Metrics Analysis For Thesis"
output: html_document
date: "2025-02-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install Packages

```{r}
library(dplyr)
library(stringr)
library(caret)
library(boot)
library(DescTools)
```

## Load all Data

```{r}
# Total manually chart reviewed cohort (n = 321)
pdr <- read.csv("/Users/ritikabatte02/Downloads/demographic_table1.csv")
# Incident cohort (n = 51)
pdr_incident <- read.csv("/Users/ritikabatte02/Downloads/DT_incidentPDR.csv")
# First ICD Method no lookback period (n = 193)
index <- read.csv("/Users/ritikabatte02/Downloads/match_idx.csv")
# Second ICD Method 1-year lookback in any department (n = 192)
anydept <- read.csv("/Users/ritikabatte02/Downloads/matching_any_dept.csv")
# Third ICD Method 1-year lookback in any ophth department (n = 158)
ophth <- read.csv("/Users/ritikabatte02/Downloads/matching_ophth.csv")
# Final summary of all 6 methods with coinciding dates
# Gold standard incident date - dr_indicator = 1, pdr_npdr = 1, pdr_date = 1
# Gold standard PDR diagnosi - dr_indicator = 1, pdr_npdr = 1
pdr_total <- read.csv("/Users/ritikabatte02/Downloads/PDR_allmethod_dates_June.csv")
```

## Remaining Data for Table 1 (n=270)

```{r}
pdr_remaining <- pdr[!pdr$PatientDurableKey %in% pdr_incident$PatientDurableKey, ]
dim(pdr_remaining)
```

## Table 1: Demographics

## View Total Cohort Data (n=321)

```{r}
head(pdr)
```

## Features for Table 1 [Total Cohort]

# Categorize by Gender

```{r}
pdr$SexBinary <- ifelse(pdr$Sex == 'Female', 1, 0)
mean_fem <- mean(pdr$SexBinary, na.rm = TRUE)
mean_fem
sd_fem <- sd(pdr$SexBinary, na.rm = TRUE)
sd_fem
prop.table(table(pdr$Sex))
```

# Categorize by PDR vs NPDR

```{r}
prop.table(table(pdr$pdr_npdr)*100)
table(pdr$pdr_npdr)
```

# Categorize by PDR vs NPDR

```{r}
prop.table(table(pdr$pdr_npdr)*100)
table(pdr$pdr_npdr)
```

# Categorize by DR diagnosis

```{r}
prop.table(table(pdr$dr_indicator))
```

# Mean and SD of Age

```{r}
mean(pdr$Age)
sd(pdr$Age)
```

# Categorize by Race/Ethnicity

```{r}
race_pop <- pdr %>%
  count(UCSFDerivedRaceEthnicity_X) %>%
  mutate(proportion = n/sum(n)*100)

race_pop
```

## View Incident PDR Data (n=51)

## Features for Table 1 [Incident Cohort]

```{r}
head(pdr_incident)
dim(pdr_incident)
```

# Categorize by Gender

```{r}
pdr_incident$SexBinary <- ifelse(pdr_incident$Sex == 'Female', 1, 0)
mean_fem <- mean(pdr_incident$SexBinary, na.rm = TRUE)
mean_fem
sd_fem <- sd(pdr_incident$SexBinary, na.rm = TRUE)
sd_fem
prop.table(table(pdr_incident$SexBinary))
```

# Categorize by PDR vs NPDR

```{r}
prop.table(table(pdr_incident$pdr_npdr)*100)
table(pdr_incident$pdr_npdr)
```

# Mean and SD of Age

```{r}
mean(pdr_incident$Age)
sd(pdr_incident$Age)
```

# Categorize by Race/Ethnicity

```{r}
race_pop <- pdr_incident %>%
  count(UCSFDerivedRaceEthnicity_X) %>%
  mutate(proportion = n/sum(n)*100)
race_pop
```

## View Remaining Cohort Data (n = 270)

```{r}
head(pdr_remaining)
dim(pdr_remaining)
```

## Features for Table 1 [Remaining Cohort]

# Categorize by Gender

```{r}
pdr_remaining$SexBinary <- ifelse(pdr_remaining$Sex == 'Female', 1, 0)
mean_fem <- mean(pdr_remaining$SexBinary, na.rm = TRUE)
mean_fem
sd_fem <- sd(pdr_remaining$SexBinary, na.rm = TRUE)
sd_fem
```

# Categorize by PDR vs NPDR

```{r}
prop.table(table(pdr_remaining$pdr_npdr)*100)
table(pdr_remaining$pdr_npdr)
```

# Mean and SD of Age

```{r}
mean(pdr_remaining$Age)
sd(pdr_remaining$Age)
```

# Categorize by Race/Ethnicity

```{r}
race_pop <- pdr_remaining %>%
  count(UCSFDerivedRaceEthnicity_X) %>%
  mutate(proportion = n/sum(n)*100)

race_pop
```

## Finding incident cases across all methods

# ICD 1 have dates

```{r}
icd1_date <- index %>%
  filter(
    !is.na(IndexDate))
nrow(icd1_date)  
```

# ICD 2 have dates

```{r}
icd2_date <- anydept %>%
  filter(!is.na(IndexDate))
nrow(icd2_date)  
```

# ICD 3 have dates

```{r}
icd3_date <- ophth %>%
  filter(
         !is.na(IndexDate))
nrow(icd3_date)  
```

# NLP have dates

```{r}
nlp_dates <- pdr_total %>%
  filter(NLP_pdr == 1,
         !is.na(NLP_date))
nrow(nlp_dates)
```

# Combo have dates

```{r}
# Create a column for ophth 
pdr_total$ophth <- ifelse(
  pdr_total$PatientDurableKey %in% ophth$PatientDurableKey,
  1, 0
)

combo_dates <- pdr_total %>%
  filter((ophth == 1 | NLP_pdr == 1), 
         !is.na(ICD_Index_Date),
         !is.na(NLP_date))
nrow(combo_dates)
```

# LLM have dates

```{r}
llm_dates <- pdr_total %>%
  filter(LLM_pdr == 1,
         !is.na(LLM_date))
nrow(llm_dates)
```

# GS date

```{r}
gs_date <- pdr_total %>%
  filter(
         pdr_npdr == 1, 
         pdr_date ==1 
)
nrow(gs_date)
```

## Finding incident PDR to the GS

# ICD 9/10 code with no lookback period

```{r}
head(index)
dim(index)
index_incident <- index %>%
  filter(dr_indicator == 1, 
         pdr_npdr == 1, 
         pdr_date == 1, 
         !is.na(IndexDate))
nrow(index_incident)
```

# Match incident ICD no lookback period to incident GS date

```{r}
icd1_date_match <- index_incident %>%
  filter(abs(as.numeric(difftime(IndexDate, pdr_date_n, units = "days"))) <= 90)

total <- nrow(index_incident)
match <- nrow(icd1_date_match)
percent_match <- (match / total) *100
print(percent_match)
```

# ICD 9/10 code with 1 yr

```{r}
head(anydept)
dim(anydept)
anydept_incident <- anydept %>%
  filter(dr_indicator == 1, 
         pdr_npdr == 1, 
         pdr_date == 1, 
         !is.na(IndexDate))
nrow(anydept_incident)
```

# Match incident ICD all department to incident GS date

```{r}
icd2_date_match <- anydept_incident %>%
  filter(abs(as.numeric(difftime(IndexDate, pdr_date_n, units = "days"))) <= 90)

total2 <- nrow(anydept_incident)
match2 <- nrow(icd2_date_match)
percent_match2 <- (match2 / total2) *100
print(percent_match2)
```

# ICD 9/10 code with 1 yr ophth

```{r}
head(ophth)
dim(ophth)
ophth_incident <- ophth %>%
  filter(dr_indicator == 1, 
         pdr_npdr == 1, 
         pdr_date == 1, 
         !is.na(IndexDate))
nrow(ophth_incident)
```

# Match incident ICD ophth department to incident GS date

```{r}
icd3_date_match <- ophth_incident %>%
  filter(abs(as.numeric(difftime(IndexDate, pdr_date_n, units = "days"))) <= 90)

total3 <- nrow(ophth_incident)
match3 <- nrow(icd3_date_match)
percent_match3 <- (match3 / total3) *100
print(percent_match3)
```

# NLP method

```{r}
nlp_incident <- pdr_total %>%
  filter(dr_indicator == 1, 
         pdr_npdr == 1, 
         pdr_date == 1,
         NLP_pdr == 1,
         !is.na(NLP_date))
nrow(nlp_incident)
```

# Match incident NLP_date to incident GS date

```{r}
nlp_date_match <- nlp_incident %>%
  filter(abs(as.numeric(difftime(NLP_date, pdr_date_n, units = "days"))) <= 90)

total <- nrow(nlp_incident)
match <- nrow(nlp_date_match)
percent_match <- (match / total) *100
print(percent_match)
```

# LLM method

```{r}
llm_incident <- pdr_total %>%
  filter(dr_indicator == 1, 
         pdr_npdr == 1, 
         pdr_date == 1,
         LLM_pdr == 1, 
         !is.na(LLM_date))
nrow(llm_incident)
```

# Match incident LLM_date to incident GS date

```{r}
llm_date_match <- llm_incident %>%
  filter(abs(as.numeric(difftime(LLM_date, pdr_date_n, units = "days"))) <= 90)

total <- nrow(llm_incident)
match <- nrow(llm_date_match)
percent_match <- (match / total) *100
print(percent_match)
```

# Combo method

```{r}
combo_NONincident <- pdr_total %>%
  filter(NLP_pdr == 1 |
         ophth == 1)
nrow(combo_NONincident)
```

```{r}
combo_incident <- pdr_total %>%
  filter(dr_indicator == 1, 
         pdr_npdr == 1, 
         pdr_date == 1,
         (NLP_pdr == 1 | ophth == 1),
         !is.na(NLP_date),
         !is.na(ICD_Index_Date)
)
nrow(combo_incident)
```

# Match incident date between NLP_date and ICD3 date to the GS date for Combination Method

```{r}
combo_date_match <- combo_incident %>%
  filter(abs(as.numeric(difftime(NLP_date, pdr_date_n, units = "days"))) <= 90 |
         abs(as.numeric(difftime(ICD_Index_Date, pdr_date_n, units = "days"))) <= 90)

total <- nrow(combo_incident)
match <- nrow(combo_date_match)
percent_match <- (match / total) *100
print(percent_match)
```

# Overlap: Of the 51 GS patients, how many appear in each method

```{r}
sum(icd1_date$PatientDurableKey %in% pdr_incident$PatientDurableKey)
sum(icd2_date$PatientDurableKey %in% pdr_incident$PatientDurableKey)
sum(icd3_date$PatientDurableKey %in% pdr_incident$PatientDurableKey)
sum(nlp_dates$PatientDurableKey %in% pdr_incident$PatientDurableKey)
sum(combo_dates$PatientDurableKey %in% pdr_incident$PatientDurableKey)
sum(llm_dates$PatientDurableKey %in% pdr_incident$PatientDurableKey)
```

# Seeing the overlapped patients, are they the same?

```{r}
# Example: Overlapping patients between ICD1 and gold standard
icd1_overlap <- intersect(icd1_date$PatientDurableKey, pdr_incident$PatientDurableKey)

# Repeat for others
icd2_overlap <- intersect(icd2_date$PatientDurableKey, pdr_incident$PatientDurableKey)
icd3_overlap <- intersect(icd3_date$PatientDurableKey, pdr_incident$PatientDurableKey)
nlp_overlap <- intersect(nlp_dates$PatientDurableKey, pdr_incident$PatientDurableKey)
combo_overlap <- intersect(combo_dates$PatientDurableKey, pdr_incident$PatientDurableKey)
llm_overlap <- intersect(llm_dates$PatientDurableKey, pdr_incident$PatientDurableKey)

all_ids <- unique(pdr_incident$PatientDurableKey)

overlap_df <- data.frame(
  PatientDurableKey = all_ids,
  ICD1 = all_ids %in% icd1_overlap,
  ICD2 = all_ids %in% icd2_overlap,
  ICD3 = all_ids %in% icd3_overlap,
  NLP = all_ids %in% nlp_overlap,
  Combo = all_ids %in% combo_overlap,
  LLM = all_ids %in% llm_overlap
)

print(overlap_df)
```

# How many patients and which ones

```{r}
sum(rowSums(overlap_df[, -1]) == 6)
overlap_df[rowSums(overlap_df[, -1]) == 6, ]
```

# Confusion Matrix

## Boostrap function 95% CI

```{r}
f1_boot <- function(data, indices) {
  d <- data[indices, ]
  TP <- sum(d$pred == 1 & d$ref == 1)
  FP <- sum(d$pred == 1 & d$ref == 0)
  FN <- sum(d$pred == 0 & d$ref == 1)
  precision <- ifelse((TP + FP) == 0, 0, TP / (TP + FP))
  recall <- ifelse((TP + FN) == 0, 0, TP / (TP + FN))
  f1 <- ifelse((precision + recall) == 0, 0, 2 * (precision * recall) / (precision + recall))
  return(f1)
}
```

# 1 ICD

```{r}
predicted <- c(rep(1, 193), rep(0, 128))  # 193 PDR, 128 no-PDR 
reference <- c(rep(1, 46),  # TP
               rep(0, 147), # FP
               rep(1, 5),  # FN
               rep(0, 123)) # TN
confusionMatrix(factor(predicted), factor(reference), mode='everything', positive = "1")

# 95% CI for proportions
TP <- 46
FP <- 147
FN <- 5
TN <- 123

P <- TP + FN
N <- TN + FP

# Sensitivity
sens_ci <- BinomCI(TP, TP + FN, method = "wilson")
# Specificity
spec_ci <- BinomCI(TN, TN + FP, method = "wilson")
# PPV 
ppv_ci <- BinomCI(TP, TP + FP, method = "wilson")
# NPV
npv_ci <- BinomCI(TN, TN + FN, method = "wilson")

# F1 score 95% CI for 1 ICD
data <- data.frame( pred= predicted, ref = reference)
set.seed(10)
boot_f1 <- boot(data = data, statistic = f1_boot, R = 1000)
f1_ci <- boot.ci(boot_f1, type = "perc")

summary_metrics <- data.frame(
  Metric = c("Sensitivity", "Specificity", "PPV", "NPV", "Accuracy", "F1 Score"),
  Estimate = round(c(sens_ci[1], spec_ci[1], ppv_ci[1], npv_ci[1], acc_ci[1], mean(boot_f1$t)), 3),
  CI_Lower = round(c(sens_ci[2], spec_ci[2], ppv_ci[2], npv_ci[2], acc_ci[2], f1_ci$percent[4]), 3),
  CI_Upper = round(c(sens_ci[3], spec_ci[3], ppv_ci[3], npv_ci[3], acc_ci[3], f1_ci$percent[5]), 3)
)

summary_metrics
```

# 2 ICD

```{r}
predicted <- c(rep(1, 192), rep(0, 129))  # 192 PDR, 129 no-PDR 
reference <- c(rep(1, 45),  # TP
               rep(0, 147), # FP
               rep(1, 6),  # FN
               rep(0, 123)) # TN
confusionMatrix(factor(predicted), factor(reference), mode='everything', positive = "1")

# 95% CI for proportions
TP <- 45
FP <- 147
FN <- 6
TN <- 123

P <- TP + FN
N <- TN + FP

# Sensitivity
sens_ci <- BinomCI(TP, TP + FN, method = "wilson")
# Specificity
spec_ci <- BinomCI(TN, TN + FP, method = "wilson")
# PPV 
ppv_ci <- BinomCI(TP, TP + FP, method = "wilson")
# NPV
npv_ci <- BinomCI(TN, TN + FN, method = "wilson")

# F1 score 95% CI for 1 ICD
data <- data.frame( pred= predicted, ref = reference)
set.seed(10)
boot_f1 <- boot(data = data, statistic = f1_boot, R = 1000)
f1_ci <- boot.ci(boot_f1, type = "perc")

summary_metrics <- data.frame(
  Metric = c("Sensitivity", "Specificity", "PPV", "NPV", "Accuracy", "F1 Score"),
  Estimate = round(c(sens_ci[1], spec_ci[1], ppv_ci[1], npv_ci[1], acc_ci[1], mean(boot_f1$t)), 3),
  CI_Lower = round(c(sens_ci[2], spec_ci[2], ppv_ci[2], npv_ci[2], acc_ci[2], f1_ci$percent[4]), 3),
  CI_Upper = round(c(sens_ci[3], spec_ci[3], ppv_ci[3], npv_ci[3], acc_ci[3], f1_ci$percent[5]), 3)
)

summary_metrics
```

# 3 ICD

```{r}
predicted <- c(rep(1, 158), rep(0, 163))  # 158 PDR, 163 no-PDR 
reference <- c(rep(1, 37),  # TP
               rep(0, 121), # FP
               rep(1, 14),  # FN
               rep(0, 149)) # TN
confusionMatrix(factor(predicted), factor(reference), mode='everything', positive = "1")

# 95% CI for proportions
TP <- 37
FP <- 121
FN <- 14
TN <- 149

P <- TP + FN
N <- TN + FP

# Sensitivity
sens_ci <- BinomCI(TP, TP + FN, method = "wilson")
# Specificity
spec_ci <- BinomCI(TN, TN + FP, method = "wilson")
# PPV 
ppv_ci <- BinomCI(TP, TP + FP, method = "wilson")
# NPV
npv_ci <- BinomCI(TN, TN + FN, method = "wilson")

# F1 score 95% CI for 1 ICD
data <- data.frame( pred= predicted, ref = reference)
set.seed(10)
boot_f1 <- boot(data = data, statistic = f1_boot, R = 1000)
f1_ci <- boot.ci(boot_f1, type = "perc")

summary_metrics <- data.frame(
  Metric = c("Sensitivity", "Specificity", "PPV", "NPV", "Accuracy", "F1 Score"),
  Estimate = round(c(sens_ci[1], spec_ci[1], ppv_ci[1], npv_ci[1], acc_ci[1], mean(boot_f1$t)), 3),
  CI_Lower = round(c(sens_ci[2], spec_ci[2], ppv_ci[2], npv_ci[2], acc_ci[2], f1_ci$percent[4]), 3),
  CI_Upper = round(c(sens_ci[3], spec_ci[3], ppv_ci[3], npv_ci[3], acc_ci[3], f1_ci$percent[5]), 3)
)

summary_metrics
```

# NLP

```{r}
predicted <- c(rep(1, 168), rep(0, 153))  # 168 PDR, 153 no-PDR 
reference <- c(rep(1, 40),  # TP
               rep(0, 128), # FP
               rep(1, 11),  # FN
               rep(0, 142)) # TN
confusionMatrix(factor(predicted), factor(reference), mode='everything', positive = "1")

# 95% CI for proportions
TP <- 40
FP <- 128
FN <- 11
TN <- 142

P <- TP + FN
N <- TN + FP

# Sensitivity
sens_ci <- BinomCI(TP, TP + FN, method = "wilson")
# Specificity
spec_ci <- BinomCI(TN, TN + FP, method = "wilson")
# PPV 
ppv_ci <- BinomCI(TP, TP + FP, method = "wilson")
# NPV
npv_ci <- BinomCI(TN, TN + FN, method = "wilson")

# F1 score 95% CI for 1 ICD
data <- data.frame( pred= predicted, ref = reference)
set.seed(10)
boot_f1 <- boot(data = data, statistic = f1_boot, R = 1000)
f1_ci <- boot.ci(boot_f1, type = "perc")

summary_metrics <- data.frame(
  Metric = c("Sensitivity", "Specificity", "PPV", "NPV", "Accuracy", "F1 Score"),
  Estimate = round(c(sens_ci[1], spec_ci[1], ppv_ci[1], npv_ci[1], acc_ci[1], mean(boot_f1$t)), 3),
  CI_Lower = round(c(sens_ci[2], spec_ci[2], ppv_ci[2], npv_ci[2], acc_ci[2], f1_ci$percent[4]), 3),
  CI_Upper = round(c(sens_ci[3], spec_ci[3], ppv_ci[3], npv_ci[3], acc_ci[3], f1_ci$percent[5]), 3)
)

summary_metrics
```

# LLM

```{r}
predicted <- c(rep(1, 195), rep(0, 126))  # 195 PDR, 126 no-PDR 
reference <- c(rep(1, 42),  # TP
               rep(0, 153), # FP
               rep(1, 9),  # FN
               rep(0, 117)) # TN
confusionMatrix(factor(predicted), factor(reference), mode='everything', positive = "1")

# 95% CI for proportions
TP <- 42
FP <- 153
FN <- 9
TN <- 117

P <- TP + FN
N <- TN + FP

# Sensitivity
sens_ci <- BinomCI(TP, TP + FN, method = "wilson")
# Specificity
spec_ci <- BinomCI(TN, TN + FP, method = "wilson")
# PPV 
ppv_ci <- BinomCI(TP, TP + FP, method = "wilson")
# NPV
npv_ci <- BinomCI(TN, TN + FN, method = "wilson")

# F1 score 95% CI for 1 ICD
data <- data.frame( pred= predicted, ref = reference)
set.seed(10)
boot_f1 <- boot(data = data, statistic = f1_boot, R = 1000)
f1_ci <- boot.ci(boot_f1, type = "perc")

summary_metrics <- data.frame(
  Metric = c("Sensitivity", "Specificity", "PPV", "NPV", "Accuracy", "F1 Score"),
  Estimate = round(c(sens_ci[1], spec_ci[1], ppv_ci[1], npv_ci[1], acc_ci[1], mean(boot_f1$t)), 3),
  CI_Lower = round(c(sens_ci[2], spec_ci[2], ppv_ci[2], npv_ci[2], acc_ci[2], f1_ci$percent[4]), 3),
  CI_Upper = round(c(sens_ci[3], spec_ci[3], ppv_ci[3], npv_ci[3], acc_ci[3], f1_ci$percent[5]), 3)
)

summary_metrics
```

# Combo

```{r}
predicted <- c(rep(1, 201), rep(0, 120))  # 201 PDR, 120 no-PDR 
reference <- c(rep(1, 38),  # TP
               rep(0, 163), # FP
               rep(1, 13),  # FN
               rep(0, 107)) # TN
confusionMatrix(factor(predicted), factor(reference), mode='everything', positive = "1")

# 95% CI for proportions
TP <- 38
FP <- 163
FN <- 13
TN <- 107

P <- TP + FN
N <- TN + FP

# Sensitivity
sens_ci <- BinomCI(TP, TP + FN, method = "wilson")
# Specificity
spec_ci <- BinomCI(TN, TN + FP, method = "wilson")
# PPV 
ppv_ci <- BinomCI(TP, TP + FP, method = "wilson")
# NPV
npv_ci <- BinomCI(TN, TN + FN, method = "wilson")

# F1 score 95% CI for 1 ICD
data <- data.frame( pred= predicted, ref = reference)
set.seed(10)
boot_f1 <- boot(data = data, statistic = f1_boot, R = 1000)
f1_ci <- boot.ci(boot_f1, type = "perc")

summary_metrics <- data.frame(
  Metric = c("Sensitivity", "Specificity", "PPV", "NPV", "Accuracy", "F1 Score"),
  Estimate = round(c(sens_ci[1], spec_ci[1], ppv_ci[1], npv_ci[1], acc_ci[1], mean(boot_f1$t)), 3),
  CI_Lower = round(c(sens_ci[2], spec_ci[2], ppv_ci[2], npv_ci[2], acc_ci[2], f1_ci$percent[4]), 3),
  CI_Upper = round(c(sens_ci[3], spec_ci[3], ppv_ci[3], npv_ci[3], acc_ci[3], f1_ci$percent[5]), 3)
)

summary_metrics
```
